Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-02-10 response-to-reviews.Rmd

I have been really interested in the comparison between gradients in canopy of individual trees and canopy made of an assemblage of species (paragraph 3.1 and 3.2. However, the paragraph 3.2 is relatively short, and I am curious if it can be developed a little bit more (maybe not!).

*[issue #99](https://github.com/EcoClimLab/vertical-thermal-review/issues/99)*

I find it interesting to try and link the empirical and theoretical knowledge on vertical gradients in forests with what is done in DGCMs and what should be done to improve their accuracy. However, I think there is a big step between the review (part I and II) and the author’s view on the implications for models (Scaling across space and time). I wonder if it would be possible to link a little bit more the different parts, and maybe without going that much in the technical details (cohort, big leaf, multi layers, ...) which I think are sometimes imprecise or maybe wrong. Would it be possible to explain more simply what are the vertical gradients presented in the section II, that are misrepresented today in DGCMs? Maybe by explaining which of the ‘biophysical variables’ and ‘leaf traits’ that you listed, are considered constant when they should be variable vertically?

**Thank you for the comment. We included a section that explains how vertical gradients are represented in models:**

- **Models partition incoming radiation and radiation within the forest vertical profile (i.e., direct vs. diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018a]. Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., CLM), while models with multiple vegetative layers can analytically solve the 2-stream approximation for each layer. Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing LAI, or lower light conditions [e.g., @krinnerDynamicGlobalVegetation2005a].**

**Given that models vary greatly in how they capture vertical light gradients, what biophysical variables are included and how those variables are parameterized to vary (or not) across vertical gradients, we feel that it is beyond the scope of this paper to comprehensively list or explain which of those variables vary vertically across different models.**

### 2022-02-10 response-to-reviews.Rmd

L429 I am not fully sure if I understood this sentence, consider maybe developing the idea. Do you mean that the variation is higher in canopy made of multi species than on the canopy of single species? If so, I didn’t understand the comparison with understory species.

**We removed this sentence because we agreed that it was confusing, and the points addressed from comments above provide substantial information for inter- and intra-canopy leaf traits.**

L470 A thought: Stomatal conductance is highly correlated to photosynthetic capacity (Wong et al., 1979) so the gradients in conductance are expected to follow the gradients in photosynthetic capacity. The water use efficiency (gs/A) or the slope parameter of conductance models (Medlyn et al., 2011) could change vertically inside the profiles. I don’t know if there are a lot of studies on the vertical variation of the water use efficiency or the slope parameter, but that could be something interesting to mention. See for example (Buckley, 2021).

**We appreciate this comment, and have added mention of the fact that maximum $g_s$ increases with light in the canopy because it typically tracks photosynthetic capacity, per @wongStomatalConductanceCorrelates1979. There is a great deal of literature showing that the relative limitation of photosynthesis by stomata (gs/A, or intrinsic WUE) tends to be greater in the upper canopy, which results in lower intercellular CO~2~ and C isotope discrimination, as noted in the original manuscript (lines 478).**

@@

More details on my thoughts for the ‘Scaling across space and time’ paragraph:

Most DGCMs represent a vertical variability in the main physiological parameters (Vcmax, Jmax, Rdark, Na). See for example (Krinner et al., 2005; Clark et al., 2011; Oleson et al., 2013). They all use some form of an exponential decrease in the photosynthetic parameters from the canopy to the ground. Since a lot of parameters and variables depend on Vcmax (for example A and therefore gs), this allows the representation of the vertical variation of a lot of leaf traits and variables. Big leaf models have this capability (see for example Krinner et al. 2005 that you cited, Appendix, or Clark et al. 2011). I think it is not clear in the way you wrote this paragraph, and it looks like ‘big leaf’ models do not represent any vertical variation at all. Note also that it is possible to use multi-layer models with or without considering shaded and sunlit leaves (See for example Clark et al. 2011).

**Thank you for noting the need to clarify this point. We have added a passage at the end of the sentence that mentions big-leaf models to clarify that they do indeed account for assumed vertical profiles: “...to big-leaf models that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer, ***implicitly capturing vertical profiles in light, photosynthetic capacity and other features by assuming those profiles are exponential and thus can be integrated analytically…”***(added text italicized here). We also deleted the clause that occurred two sentences later, that had said about big-leaf models: “...which do not represent any vertical stratification…”**

**In addition, we also added a paragraph that more clearly describes the methods used for light competition that enables the type of variation in physiological traits that you described (e.g., Vcmax).**

- **Owing to differences in the representation of forest vertical strata, DVGMs vary in their ability to incorporate vertical variation in leaf traits and physiological processes. In general, however, this variation is accounted for via light competition. Models partition incoming radiation and radiation within the forest vertical profile (i.e., direct vs. diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018a]. Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., CLM), while models with multiple vegetative layers can analytically solve the 2-stream approximation for each layer. Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing LAI, or lower light conditions [e.g., @krinnerDynamicGlobalVegetation2005a].**

### 2022-02-10 response-to-reviews.Rmd

More details on my thoughts for the ‘Scaling across space and time’ paragraph:

Most DGCMs represent a vertical variability in the main physiological parameters (Vcmax, Jmax, Rdark, Na). See for example (Krinner et al., 2005; Clark et al., 2011; Oleson et al., 2013). They all use some form of an exponential decrease in the photosynthetic parameters from the canopy to the ground. Since a lot of parameters and variables depend on Vcmax (for example A and therefore gs), this allows the representation of the vertical variation of a lot of leaf traits and variables. Big leaf models have this capability (see for example Krinner et al. 2005 that you cited, Appendix, or Clark et al. 2011). I think it is not clear in the way you wrote this paragraph, and it looks like ‘big leaf’ models do not represent any vertical variation at all. Note also that it is possible to use multi-layer models with or without considering shaded and sunlit leaves (See for example Clark et al. 2011).

**Thank you for noting the need to clarify this point. We have added a passage at the end of the sentence that mentions big-leaf models to clarify that they do indeed account for assumed vertical profiles: “...to big-leaf models that reduce 3D vegetation structure across the entire biosphere into a single vegetation layer, ***implicitly capturing vertical profiles in light, photosynthetic capacity and other features by assuming those profiles are exponential and thus can be integrated analytically…”***(added text italicized here). We also deleted the clause that occurred two sentences later, that had said about big-leaf models: “...which do not represent any vertical stratification…”**

**In addition, we also added a paragraph that more clearly describes the methods used for light competition that enables the type of variation in physiological traits that you described (e.g., Vcmax):**

@@

- **Owing to differences in the representation of forest vertical strata, DVGMs vary in their ability to incorporate vertical variation in leaf traits and physiological processes. In general, however, this variation is accounted for via light competition. Models partition incoming radiation and radiation within the forest vertical profile (i.e., direct vs. diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018a]. Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., CLM), while models with multiple vegetative layers can analytically solve the 2-stream approximation for each layer. Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing LAI, or lower light conditions [e.g., @krinnerDynamicGlobalVegetation2005a].**

To me, an important remark is that most gradients in DGVMs are prescribed. More data would of course help to better understand and quantify the gradients, and to compare them with model representations (part ‘Scaling in situ data with remote sensing’ of your paper). But in the end, the mechanisms explaining those gradients are still not really known, in the sense that they can’t be efficiently modeled prognostically. This is due to the multiple sources of parameters variation that you highlighted (light environment, temperature, hydraulic, …). A lot of research is done using optimality frameworks to try and predict the photosynthetic and other vegetation traits based on environmental variables. See for example (Ali et al., 2016; Buckley, 2021). A question I have, is if including more precise cohorts as you suggest, would improve the accuracy of models given all the uncertainty in the parametrization of their traits.

**We agree that there is a recognition in the modeling community that we do need improved understanding of the mechanisms to better represent the level of vertical variation desired by some. We added the following sentence to acknowledge that: "However, doing so will require improved understanding of the mechanisms controlling vertical gradients.". We also removed some of the language that emphasizes cohort based models as the only or best way to do this.**

### 2022-02-10 response-to-reviews.Rmd

Certainly, a discussion with a micrometeorologist or atmospheric scientist, if not already had, could improve the manuscript. Statements like l. 156: 'Wind speeds are also higher at the top of the canopy, owing to the buffering effect of the canopy' are an awkward read. Buffering? There could be much improvement by looking at it the other way: wind will blow as it does at the top of the canopy until encountering the plant canopy top as an aerodynamic drag element; then the additional leaf area entrains air movement and sweep-eject motion and eddies are attenuated through the canopy resulting in progressively lower windspeeds deeper into canopies. I don’t know how buffering enters into this, or what the authors mean here.

**We appreciate these comments and have added a discussion of canopy transport from a more theoretical perspective (lines xxx-xxx), immediately after the paragraph about canopy radiation profiles and before those on CO~2~, H~2~O and temperature profiles. We have also modified the subsequent paragraph about wind speeds to eliminate the use of the confusing term “buffering”.**

**We had also used the words “buffering”, "buffer", or "buffered" in a different context, where the intention was to describe how canopy attenuation (of radiation and wind) can reduce daily extremes of temperature in the lower canopy and understory. We have modified the text in those places to clarify, generally reducing use of the terms. However, the term has been widely used in recent publications on this theme, and we therefore did not feel it was necessary to eliminate it's usage. Examples of modified sentences include the following:**

- **“Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions ***from large fluctuations in conditions above the canopy”; (line xxx)***(new text italicized)**
- **“Air temperature ($T_{air}$) often shows little variation across the vertical gradient, but under certain circumstances, ***attenuation of radiation and vertical transport of sensible heat by the canopy can buffer the lower canopy and understory from large diurnal swings in air temperature***~~it can be significantly buffered by forest canopies~~"; (line xxx)**
- **“Typically, ***diurnal temperature range is smaller beneath dense canopies than above, resulting in,***~~dense canopies buffer understories from high maximum $T_{air}$ more than open canopies, i.e., dense canopy understories can have~~cooler maximum daytime $T_{air}$ and warmer nighttime minimum Tair than open canopy understories or nearby clearings.” (line xxx)**

@@

**We agree, and have modified the text in this section to emphasize the role of VPD in driving stomatal closure in sun leaves (new text italicized): “However, ~~because~~ water supply often cannot meet the demands incurred by the high irradiance and $g_b$ experienced by sun leaves ***with fully open stomata, because of height-related constraints on water transport (e.g., Yoder et al. 1994, Koch et al. 2004, Sillett et al 2010) and/or because of increased evaporative demand (VPD) caused by leaf warming in sunlit canopy locations (Buckley et al. 2014)***. As a result, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table 2), which drives the lower intracellular CO2 and carbon isotope discrimination discussed in section 3.1 (Table 1).  Stomatal depression reduces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads ~~In sun leaves, Tleaf thus further increases due to the lack of transpirational cooling.~~” We have also added the phrase ***“(driven by increased VPD)”***after “sun leaves in the upper canopy show a stronger decrease in gs in response to rising Tair”, and clarified that “high Tair should decrease gs of ***upper canopy***leaves”.**

line510-525: There is a long discussion about whether Topt for gas exchange or its components differs between overstory and understory. Here the authors aren’t clear, but such a comparison only matters if the same species is measured in different canopy layers.

**Thank you for noting this, we revised the sentence to make it clear, where we specify that majority of the studies are within species between sun and shade leaves of the same trees. @carterExperimentalWarmingTropical2021 is one of the only studies comparing between overstory and understory.**

- **We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves, and existing studies, which  compare sun and shade leaves of the same trees, reveal no pronounced overall trend with height in the optimum temperatures for photosynthetic processes**

### 2022-02-10 vertical-thermal-review.Rmd

Leaf traits shape $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a;  @zwienieckiHydraulicLimitationsImposed2004].
Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determine the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016]. 
This dichotomy in leaves is observed along the vertical gradient where upper (or outer) canopy leaves vary dramatically from canopy-interior and understory leaves due to the differences in biophysical conditions [Fig. `r fig_NEON_vertical`, @binLeafTraitExpression2022], and also because of differences in biological conditions such as tree size and height [@binLeafTraitExpression2022].

>>>>

Leaf traits shape $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a;  @zwienieckiHydraulicLimitationsImposed2004].
Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determine the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016]. 
Therefore, a continuum of sun and shade leaves are observed throughout the vertical gradient where upper (or outer) canopy leaves vary dramatically from canopy-interior and understory leaves due to the differences in biophysical conditions [Fig. `r fig_NEON_vertical`, @binLeafTraitExpression2022], and also because of differences in biological conditions such as tree size and height [@binLeafTraitExpression2022].
